Effective multi-body SU($N$)-symmetric interactions of ultracold fermionic atoms on a 3-D lattice

### 2018-10-13 multi_body_interactions/papers/theory/paper.tex
minor changes to multi-body text before arXiv upload

\begin{abstract}
  Rapid advancements in the experimental capabilities with ultracold
  alkaline-earth-like atoms (AEAs) bring to a surprisingly near term
  the prospect of performing quantum simulations of spin models and
  lattice field theories exhibiting SU($N$) symmetry.  Motivated in
  particular by recent experiments preparing high density samples of
  strongly interacting ${}^{87}$Sr atoms in a three-dimensional
  optical lattice, we develop a low-energy effective theory of
  fermionic AEAs which exhibits emergent SU($N$)-symmetric multi-body
  interactions, where $N$ is the number of atomic nuclear spin levels.
  We fully characterize the low-lying excitations in our effective
  theory, and compare predictions of many-body interaction energies
  with direct measurements of many-body excitation spectra in an
  optical lattice clock.  Our work makes the first step in enabling a
  controlled, bottom-up experimental investigation of multi-body
  SU($N$) physics.
\end{abstract}

@@

Fermionic alkaline-earth atoms (AEAs), in addition to other atoms such
as ytterbium (Yb) sharing similar electronic structure, are currently
the building blocks of the most precise atomic clocks in the world
\cite{derevianko2011colloquium, katori2011optical, ludlow2015optical}.
The atoms have a unique, ultra-narrow optical transition between
metastable ${}^1S_0$ and ${}^3P_0$ electronic orbital states, i.e.~the
``clock states'', that allows for coherence times which can exceed 100
seconds \cite{porsev2004hyperfine, santra2004properties}.
Furthermore, AEAs can be trapped in fully controllable optical lattice
potentials and interrogated with ultra-stable lasers that can resolve
and probe their rich hyperfine spectra, consisting of $N$ different
nuclear spin levels with $N$ as large as 10 in strontium (${}^{87}$Sr)
and 6 in ytterbium (${}^{173}$Yb).

In 2015 the ${}^{87}$Sr optical lattice clock (OLC) at JILA, operated
in a one-dimensional (1-D) lattice at microkelvin temperatures,
achieved a total fractional uncertainty of $2\times10^{-18}$
\cite{bloom2014optical, nicholson2015systematic}.  Last year (2017), a
new generation of OLCs became operational at JILA, interrogating a
Fermi degenerate gas of ${}^{87}$Sr atoms in a 3-D lattice at
nanokelvin temperatures \cite{campbell2017fermidegenerate}.  All of
these atoms' degrees of freedom, including the electronic orbital,
nuclear spin, and motional states, can be fully controlled in a 3-D
lattice with high fidelity \cite{daley2008quantum,
  gorshkov2009alkalineearthmetal, daley2011quantum,
  daley2011statedependent}.  With frequency measurements reaching the
$10^{-19}$ fractional uncertainty level, the new OLCs are thus
enabling an exciting opportunity to probe, for the first time, quantum
dynamics with sub-millihertz spectral resolution
\cite{campbell2017fermidegenerate}.

@@

The presence of this exotic SU($N$) symmetry in a highly controllable
experimental platform opens the door to experimental studies of
e.g.~the SU($N$) Heisenberg model, whose phase diagram is believed to
exhibit features such as a chiral spin liquid (CSL) phase with
topological order and fractional statistics \cite{hermele2009mott,
  hermele2011topological, chen2016syntheticgaugefield}.  In addition
to illuminating open questions in our understanding of the fractional
quantum Hall effect and unconventional superconductivity
\cite{lee2006doping, lee2008high, gong2014emergent}, the CSL can
support non-Abelian excitations which allow for universal topological
quantum computation \cite{freedman2004class, hermele2011topological}.
Harnessing the SU($N$)-symmetric interactions of AEAs might also
enable the simulation of various lattice gauge theories
\cite{wieseu.-j.2013ultracold, zohar2016quantum}, some of which share
important qualitative features with quantum chromodynamics such as
few-body bound states and confinement \cite{banerjee2013atomic,
  rico2018nuclear}.  These direct, quantum simulations have an
extraordinary potential to provide novel insights by circumventing
severe sign problems which plague classical simulations of strongly
interacting fermionic systems \cite{wieseu.-j.2013ultracold,
  wu2003exact}.

@@

Though effective multi-body interactions have previously been studied
in the context of harmonically \cite{johnson2012effective,
  yin2014universal} and lattice-confined \cite{johnson2009effective}
neutral bosons prepared in a single hyperfine state, our work deals
for the first time with fermionic atoms with internal degrees of
freedom and multiple collisional parameters.  Nonetheless, due to the
SU($N$) symmetry of these collisions, we are able to find a simple way
to express effective multi-body Hamiltonians and fully characterize
their low-lying many-body energy eigenstates.  While some past work
has detected experimental signatures of multi-body interactions in the
form of quantum phase revivals \cite{will2010timeresolved}, we perform
a direct comparison of the many-body interaction energies predicted by
our low-energy effective theory with experimental measurements of
density-dependent atomic energy spectra performed in
ref.~\cite{goban2018emergence}, similarly to the measurements of
bosons performed in ref.~\cite{mark2011precision}.

The remainder of this paper is structured as follows.  In section
\ref{sec:overview} we provide an overview of the relevant one- and
two-body physics of ultracold atoms on a lattice, and summarize our
main result of effective multi-body interactions between these atoms.
In section \ref{sec:low_energy}, we discuss our method for deriving a
low-energy effective theory, provide a perturbative expansion for the
net effective Hamiltonian, and compute all $M$-body Hamiltonians
through third order in the low-energy effective theory.  We then
analyze the low-lying excitations of the effective theory in section
\ref{sec:spectra}, comparing spectral predictions with experimental
measurements, and study the orbital-state dynamics of nuclear spin
mixtures interrogated via Rabi spectroscopy.  Finally, we summarize
and conclude our findings in section \ref{sec:summary}, and provide
some discussion of future outlooks.

@@

Consider a collection of fermionic atoms with total nuclear spin $I$
and two atomic orbital states loaded into a translationally invariant
3-D optical lattice.  While an external trapping potential will
generally break discrete translational symmetry of the lattice, any
background inhomogeneity can be made negligible by spectroscopically
addressing a sufficiently small region of the lattice
\cite{goban2018emergence}.  Throughout this paper, we work strictly in
the deep-lattice regime with negligible tunneling between lattice
sites.  We also assume a so-called ``magic-wavelength'' lattice for
which both atomic orbital states experience identical lattice
potentials \cite{ye2008quantum}.  The single-particle Hamiltonian of
the atoms can then be written in the form
\begin{align}
  H_0 = \sum_{i,n,\mu,s} E_n \c_{in\mu s}^\dag \c_{in\mu s},
  \label{eq:H_0}
\end{align}
where $\c_{in\mu s}$ is a fermionic operator which annihilates a
single atom on lattice site $i\in\mathbb{Z}^3$ in motional state
$n\in\mathbb{N}_0^3$ with nuclear spin $\mu\in\set{-I,I+1,\cdots,I}$
(i.e.~projected onto a quantization axis) and orbital state
$s\in\set{\g,\e}$; and $E_n$ is the energy of a single atom on site
$i$ in motional state $n$.  In a harmonic trap approximation we would
have $E_n=\p{3/2+n_x+n_y+n_z}\omega$ for an on-site angular trap
frequency $\omega$ (with the reduced Planck constant $\hbar=1$
throughout this paper), but in general the aharmonicity of the lattice
potential will cause a non-negligible shift in motional state
energies.

In the absence of hyperfine coupling, as when addressing the spinless
${}^1S_0$ ($\g$) and ${}^3P_0$ ($\e$) orbital states, interactions
between any two atoms are governed by their orbital states alone, and
are therefore characterized by four scattering lengths $a_X$ with
$X\in\set{\g\g,\e\g^-,\e\g^+,\e\e}$, where the $+$ ($-$) superscript
denotes symmetrization (anti-symmetrization) of a two-body orbital
state under particle exchange.  In the low-energy limit, we can write
the bare two-body interaction Hamiltonian in the form
\cite{gorshkov2010twoorbital}
\begin{align}
  H_{\t{int}} = \sum_{\substack{\mu<\nu\\s}}
  G_s \int d^3x~ \hat\rho_{\mu s} \hat\rho_{\nu s}
  + G_+ \sum_{\mu,\nu}
  \int d^3x~ \hat\rho_{\mu,\e} \hat\rho_{\nu,\g}
  + G_- \sum_{\mu,\nu} \int d^3x~
  \hat\psi_{\mu,\e}^\dag \hat\psi_{\nu,\g}^\dag
  \hat\psi_{\nu,\e} \hat\psi_{\mu,\g},
\end{align}
where $\hat\psi_{\mu s}$ is a fermionic field operator for atoms with
nuclear spin $\mu$ and orbital state $s$;
$\hat\rho_{\mu s}\equiv\hat\psi_{\mu s}^\dag\hat\psi_{\mu s}$ is an
atomic density field operator; and the coupling constants $G_X$ are
defined in terms of the scattering lengths by
\begin{align}
  G_{s=\g,\e} \equiv \f{4\pi}{m_A}~ a_{ss}, &&
  G_\pm \equiv \f{2\pi}{m_A}~ \p{a_{\e\g+} \pm a_{\e\g-}},
  \label{eq:couplings}
\end{align}
where $m_A$ is the mass of a single atom.  Defining the coupling
tensor
\begin{align}
  G^{qr}_{st} \equiv \left\{
    \begin{array}{ll}
      G_q & ~ q = r = s = t \\
      G_+ & ~ q \ne r ~ \t{and} ~ (q,r) = (s,t) \\
      G_- & ~ q \ne r ~ \t{and} ~ (q,r) = (t,s) \\
      0 & ~ \t{otherwise}
    \end{array}\right.,
  \label{eq:coupling_tensor}
\end{align}
where $q,r,s,t\in\set{\g,\e}$ are orbital state indices, we can
alternately write the bare two-body interaction Hamiltonian in the
more compact form
\begin{align}
  H_{\t{int}}
  = \f12 \sum_{\substack{q,r,s,t\\\mu,\nu}} G^{qr}_{st} \int d^3x~
  \hat\psi_{\mu s}^\dag \hat\psi_{\nu t}^\dag
  \hat\psi_{\nu r} \hat\psi_{\mu q}.
  \label{eq:H_int_fields}
\end{align}
For nuclear spins $\mu,\nu$, the coupling tensor $G^{qr}_{st}$ gives
the coupling constant between the two-atom states
$(\mu,q)+(\nu,r)\leftrightarrow(\mu,s)+(\nu,t)$.

We now expand the field operators $\hat\psi_{\mu s}$ in the Wannier
basis for a 3-D lattice, such that
$\hat\psi_{\mu s}(x) = \sum_{i,n} \phi_{in}(x) \c_{in\mu s}$ with
spatial wavefunctions $\phi_{in}$ and fermionic annihilation operators
$\c_{in\mu s}$ indexed by lattice sites $i$ and motional states $n$.
Invoking the tight-binding approximation, we assume that the spatial
overlap integral in \eqref{eq:H_int_fields} is negligible unless all
wavefunctions are localized at the same lattice site; we discuss the
breakdown of this approximation and its consequences for our
low-energy effective theory in Appendix \ref{sec:error}.  The relevant
spatial overlap integral is then
\begin{align}
  K^{k\ell}_{mn}
  \equiv \int d^3x~ \phi_{im}^* \phi_{in}^* \phi_{i\ell} \phi_{ik},
  \label{eq:K_klmn}
\end{align}
which for a lattice with discrete translational invariance is
independent of the lattice site $i$.  The two-body interaction
Hamiltonian can be written in terms of this overlap integral as
\begin{align}
  H_{\t{int}}
  = \f12 \sum_{\substack{i,k,\ell,m,n\\q,r,s,t\\\mu,\nu}}
  K^{k\ell}_{mn} G^{qr}_{st}
  \c_{im\mu s}^\dag \c_{in\nu t}^\dag \c_{i\ell\nu r} \c_{ik\mu q}
  \equiv \f12 \sum K^{k\ell}_{mn} G^{qr}_{st}
  \c_{m\mu s}^\dag \c_{n\nu t}^\dag \c_{\ell\nu r} \c_{k\mu q},
  \label{eq:H_int}
\end{align}
where for brevity we will henceforth suppress the identical site index
($i$) on all operators, and implicitly sum over all indices in a
summand which do not have a fixed value.  We may also at times
suppress motional state indices on the overlap integral
$K^{k\ell}_{mn}$, in which case the suppressed indices are implicitly
zero (corresponding to a motional ground state); i.e.
\begin{align}
  K^{\ell m}_n \equiv K^{\ell m}_{n,0},
  &&
  K^m_n \equiv K^{m,0}_{n,0},
  &&
  K_{mn} \equiv K^{0,0}_{mn},
  &&
  K_n \equiv K^{0,0}_{n,0},
  &&
  K \equiv K^{0,0}_{0,0}.
  \label{eq:K}
\end{align}
For simplicity, we will also generally work in a gauge for which all
two-body overlap integrals are real, such that
$K^{k\ell}_{mn}={K^{k\ell}_{mn}}^*=K^{mn}_{k\ell}$.  This choice is
always possible due to the analytic properties of Wannier orbitals
$\phi_{in}$ \cite{kohn1959analytic}.

We are interested in the ground and low-lying orbital eigenstates of
several interacting atoms on a single lattice site, as such states can
be readily prepared and coherently addressed in current experiments
with ultracold atoms \cite{goban2018emergence}.  Although these
experiments operate at temperatures sufficiently low for negligible
thermal occupation of excited motional states, simply neglecting
excited single-particle motional states fails to reproduce the
observed spectra of multiply-occupied lattice sites
\cite{goban2018emergence}.  The reason for this failure lies in the
fact that the bare two-body Hamiltonian $H_{\t{int}}$ in
\eqref{eq:H_int} is not diagonal in the basis of single-particle
motional states.  Even at zero temperature, therefore, the presence of
excited motional states modifies many-body atomic energy spectra when
off-diagonal elements of $H_{\t{int}}$ approach the scale of low-lying
single-particle motional state energy gaps.  To address this problem,
we develop a low-energy effective theory for ultracold two-level
fermionic atoms on a lattice.  As our work is motivated by progress in
experiments which initialize all atoms in identical (i.e.~ground)
orbital and motional states, to simplify our theory we assume that any
$N$ atoms on a single lattice site occupy $N$ nuclear spin states.
Multiple occupation of a single nuclear spin state on the same lattice
site is initially forbidden by fermionic statistics, and subsequently
cannot be violated in the absence of tunneling and hyperfine coupling.

Our low-energy effective theory exhibits SU($N$)-symmetric multi-body
interactions, such that the effective interaction Hamiltonian can be
written in the form
\begin{align}
  H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} H_M,
\end{align}
where $H_M$ is an $M$-body Hamiltonian, $I$ is the total nuclear spin
of each atom (e.g.~$I=9/2$ for ${}^{87}$Sr), and the sum ends at
$2I+1$ because this is the largest number of atoms which may initially
occupy a single lattice site.  Focusing on the low-lying orbital state
excitations of this theory, we find that we can express multi-body
Hamiltonians in the simple form
\begin{align}
  H_M = \sum_{\abs{\set{\mu_j}}=M}
  \p{U_{M,\g} \prod_{\alpha=1}^M \n_{\mu_\alpha,\g}
    + U_{M,+} \n_{\mu_1,\e} \prod_{\alpha=2}^M \n_{\mu_\alpha,\g}
    + U_{M,-} \c_{\mu_1,\g}^\dag \c_{\mu_2,\e}^\dag
    \c_{\mu_2,\g} \c_{\mu_1,\e} \prod_{\alpha=3}^M \n_{\mu_\alpha,\g}},
  \label{eq:H_M_preview}
\end{align}
where $\c_{\mu s}\equiv\c_{0,\mu s}$ is a fermionic annihilation
operator for an atom in the motional ground state with nuclear spin
$\mu$ and orbital state $s$;
$\n_{\mu s}\equiv \c_{\mu s}^\dag \c_{\mu s}$ is a number operator;
and the sum is performed over all choices of nuclear spins $\mu_j$
with $j=1,\cdots,M$ for which all $\mu_j$ are distinct, or
equivalently all choices of $\mu_j$ for which the set $\set{\mu_j}$
contains $M$ elements, for a total of ${2I+1\choose M}\times\p{M!}$
nuclear spin combinations.  The coefficients $U_X$ can be expressed in
terms of the coupling constants of the bare two-body interaction
Hamiltonian (see Appendix \ref{sec:U_X}).  The key feature of the
$M$-body interactions in \eqref{eq:H_M_preview} is that they take a
form identical to the two-body interactions between motional
ground-state atoms, but with the addition of $M-2$ spectator atoms.
This form is a direct consequence of the SU($N$) symmetry of
underlying two-body interactions.

@@

As the bare two-body Hamiltonian $H_{\t{int}}$ is not diagonal with
respect to single-particle motional states, the problem of determining
many-body spectra and eigenstates nominally involves all motional
degrees of freedom of the atoms involved.  If we are interested in
addressing internal atomic states at zero temperature, however, in
principle we need only to consider the motional ground state
associated with each many-body internal atomic state.  We can then
ignore all excited many-body motional states, as they will be neither
thermally occupied nor externally interrogated.  This procedure makes
the many-body problem more tractable by greatly reducing the
dimensionality of our target Hilbert space.

@@

order, we introduce an expanded coupling tensor

@@

indices, the expanded coupling tensor allows us to collect together
diagrams which have identical graph topologies, but address different

@@

associated negative sign if they contain only one excited virtual
state, and a factor of $1/2$ if they contain a virtual ground state.

@@

in a contribution to the effective Hamiltonian which is equal to

@@

the second and third of these diagrams pick up a factor of $-1/2$, and
the sum of the disconnected diagrams vanishes.

>>>>

\begin{abstract}
  Rapid advancements in the experimental capabilities with ultracold
  alkaline-earth-like atoms (AEAs) bring to a surprisingly near term
  the prospect of performing quantum simulations of spin models and
  lattice field theories exhibiting SU($N$) symmetry.  Motivated in
  particular by recent experiments preparing high density samples of
  strongly interacting ${}^{87}$Sr atoms in a three-dimensional
  optical lattice, we develop a low-energy effective theory of
  fermionic AEAs which exhibits emergent multi-body SU($N$)-symmetric
  interactions, where $N$ is the number of atomic nuclear spin levels.
  We fully characterize the low-lying excitations in our effective
  theory, and compare predictions of many-body interaction energies
  with direct measurements of many-body excitation spectra in an
  optical lattice clock.  Our work makes the first step in enabling a
  controlled, bottom-up experimental investigation of multi-body
  SU($N$) physics.
\end{abstract}

@@

Fermionic alkaline-earth atoms (AEAs), in addition to other atoms such
as ytterbium (Yb) sharing similar electronic structure, are currently
the building blocks of the most precise atomic clocks in the world
\cite{derevianko2011colloquium, katori2011optical, ludlow2015optical}.
These atoms have a unique, ultra-narrow optical transition between
metastable ${}^1S_0$ and ${}^3P_0$ electronic orbital states, i.e.~the
``clock states'', that allows for coherence times which can exceed 100
seconds \cite{porsev2004hyperfine, santra2004properties}.
Furthermore, AEAs can be trapped in fully controllable optical lattice
potentials and interrogated with ultra-stable lasers that can resolve
and probe their rich hyperfine spectra, consisting of $N$ different
nuclear spin levels with $N$ as large as 10 in strontium (${}^{87}$Sr)
and 6 in ytterbium (${}^{173}$Yb).

In 2015 the ${}^{87}$Sr optical lattice clock (OLC) at JILA, operated
in a one-dimensional (1-D) lattice at microkelvin temperatures,
achieved a total fractional uncertainty of $2\times10^{-18}$
\cite{bloom2014optical, nicholson2015systematic}.  Last year (2017), a
new generation of OLCs became operational at JILA, interrogating a
Fermi degenerate gas of ${}^{87}$Sr atoms in a 3-D lattice at
nanokelvin temperatures \cite{campbell2017fermidegenerate}.  All of
these atoms' degrees of freedom, including the electronic orbital,
nuclear spin, and motional states, can be fully controlled with high
fidelity in a 3-D lattice \cite{daley2008quantum,
  gorshkov2009alkalineearthmetal, daley2011quantum,
  daley2011statedependent}.  With frequency measurements reaching the
$10^{-19}$ fractional uncertainty level, the new OLCs are thus
enabling an exciting opportunity to probe, for the first time, quantum
dynamics with sub-millihertz spectral resolution
\cite{campbell2017fermidegenerate}.

@@

The presence of this exotic SU($N$) symmetry in a highly controllable
experimental platform opens the door to experimental studies of
e.g.~the SU($N$) Heisenberg model, whose phase diagram is believed to
exhibit features such as a chiral spin liquid (CSL) phase with
topological order and fractional statistics \cite{hermele2009mott,
  hermele2011topological, chen2016syntheticgaugefield}.  In addition
to illuminating open questions in our understanding of the fractional
quantum Hall effect and unconventional superconductivity
\cite{lee2006doping, lee2008high, gong2014emergent}, the CSL can
support non-Abelian excitations which allow for universal topological
quantum computation \cite{freedman2004class, hermele2011topological}.
Harnessing the SU($N$)-symmetric interactions of AEAs might also
enable the simulation of various lattice gauge theories
\cite{wieseu.-j.2013ultracold, zohar2016quantum}, some of which share
important qualitative features with quantum chromodynamics such as
few-body bound states and confinement \cite{banerjee2013atomic,
  rico2018nuclear}.  These direct, quantum simulations have an
extraordinary potential to provide novel insights by circumventing
e.g.~severe sign problems which plague classical simulations of
strongly interacting fermionic systems \cite{wieseu.-j.2013ultracold,
  wu2003exact}.

@@

Though effective multi-body interactions have previously been studied
in the context of harmonically \cite{johnson2012effective,
  yin2014universal} and lattice-confined \cite{johnson2009effective}
neutral bosons prepared in a single hyperfine state, our work deals
for the first time with fermions that have internal degrees of freedom
and multiple collisional parameters.  Nonetheless, due to the SU($N$)
symmetry of these collisions, we are able to find a simple way to
express effective multi-body Hamiltonians and fully characterize their
low-lying many-body energy eigenstates.  While some past work has
detected experimental signatures of multi-body interactions in the
form of quantum phase revivals \cite{will2010timeresolved}, we perform
a direct comparison of the many-body interaction energies predicted by
our low-energy effective theory with experimental measurements of
density-dependent atomic energy spectra performed in
ref.~\cite{goban2018emergence}, similarly to the measurements with
bosons performed in ref.~\cite{mark2011precision}.

The remainder of this paper is structured as follows.  In section
\ref{sec:overview} we provide an overview of the relevant one- and
two-body physics of ultracold atoms on a lattice, and summarize our
main result of effective multi-body interactions between these atoms.
In section \ref{sec:low_energy} we discuss our method for deriving a
low-energy effective theory, provide a perturbative expansion for the
net effective Hamiltonian, and compute all $M$-body Hamiltonians
through third order in the low-energy effective theory.  We then
analyze the low-lying excitations of the effective theory in section
\ref{sec:spectra}, comparing spectral predictions with experimental
measurements, and study the orbital-state dynamics of nuclear spin
mixtures interrogated via Rabi spectroscopy.  Finally, we summarize
and conclude our findings in section \ref{sec:summary}, and provide
some discussion of future outlooks.

@@

Consider a collection of fermionic atoms with total nuclear spin $I$
and two atomic orbital states loaded into a translationally invariant
3-D optical lattice.  While an external trapping potential will
generally break discrete translational symmetry of the lattice, any
background inhomogeneity can be made negligible by spectroscopically
addressing a sufficiently small region of the lattice
\cite{goban2018emergence}.  Throughout this paper, we work strictly in
the deep-lattice regime with negligible tunneling between lattice
sites.  We also assume a so-called ``magic-wavelength'' lattice for
which both atomic orbital states experience identical lattice
potentials \cite{ye2008quantum}.  The single-particle Hamiltonian of
the atoms can then be written in the form
\begin{align}
  H_0 = \sum_{i,n,\mu,s} E_n \c_{in\mu s}^\dag \c_{in\mu s},
  \label{eq:H_0}
\end{align}
where $\c_{in\mu s}$ is a fermionic operator which annihilates a
single atom on lattice site $i\in\mathbb{Z}^3$ in motional state
$n\in\mathbb{N}_0^3$ with nuclear spin $\mu\in\set{-I,I+1,\cdots,I}$
(i.e.~projected onto a quantization axis) and orbital state
$s\in\set{\g,\e}$; and $E_n$ is the energy of a single atom in
motional state $n$.  In a harmonic trap approximation we would have
$E_n=\p{3/2+n_x+n_y+n_z}\omega$ for an on-site angular trap frequency
$\omega$ (with the reduced Planck constant $\hbar=1$ throughout this
paper), but in general the aharmonicity of the lattice potential will
cause a non-negligible shift in motional state energies.

In the absence of hyperfine coupling, as when addressing the spinless
${}^1S_0$ ($\g$) and ${}^3P_0$ ($\e$) orbital states, interactions
between any two atoms are governed by their orbital states alone, and
are therefore characterized by four scattering lengths $a_X$ with
$X\in\set{\g\g,\e\g^-,\e\g^+,\e\e}$, where the $+$ ($-$) superscript
denotes symmetrization (anti-symmetrization) of a two-body orbital
state under particle exchange.  In the low-energy limit, we can write
the bare two-body interaction Hamiltonian in the form
\cite{gorshkov2010twoorbital}
\begin{align}
  H_{\t{int}} = \sum_{\substack{\mu<\nu\\s}}
  G_s \int d^3x~ \hat\rho_{\mu s} \hat\rho_{\nu s}
  + G_+ \sum_{\mu,\nu}
  \int d^3x~ \hat\rho_{\mu,\e} \hat\rho_{\nu,\g}
  + G_- \sum_{\mu,\nu} \int d^3x~
  \hat\psi_{\mu,\e}^\dag \hat\psi_{\nu,\g}^\dag
  \hat\psi_{\nu,\e} \hat\psi_{\mu,\g},
\end{align}
where $\hat\psi_{\mu s}$ is a fermionic field operator for atoms with
nuclear spin $\mu$ and orbital state $s$;
$\hat\rho_{\mu s}\equiv\hat\psi_{\mu s}^\dag\hat\psi_{\mu s}$ is an
atomic density field operator; and the coupling constants $G_X$ are
defined in terms of the scattering lengths by
\begin{align}
  G_{s=\g,\e} \equiv \f{4\pi}{m_A}~ a_{ss}, &&
  G_\pm \equiv \f{2\pi}{m_A}~ \p{a_{\e\g+} \pm a_{\e\g-}},
  \label{eq:couplings}
\end{align}
where $m_A$ is the mass of a single atom.  Defining for brevity
\begin{align}
  G^{qr}_{st} \equiv \left\{
    \begin{array}{ll}
      G_q & ~ q = r = s = t \\
      G_+ & ~ q \ne r ~ \t{and} ~ (q,r) = (s,t) \\
      G_- & ~ q \ne r ~ \t{and} ~ (q,r) = (t,s) \\
      0 & ~ \t{otherwise}
    \end{array}\right.,
  \label{eq:coupling_tensor}
\end{align}
where $q,r,s,t\in\set{\g,\e}$ are orbital state indices, we can
alternately write the bare two-body interaction Hamiltonian in the
more compact form
\begin{align}
  H_{\t{int}}
  = \f12 \sum_{\substack{q,r,s,t\\\mu,\nu}} G^{qr}_{st} \int d^3x~
  \hat\psi_{\mu s}^\dag \hat\psi_{\nu t}^\dag
  \hat\psi_{\nu r} \hat\psi_{\mu q}.
  \label{eq:H_int_fields}
\end{align}
For nuclear spins $\mu,\nu$, the symbol $G^{qr}_{st}$ gives the
coupling constant between the two-atom states
$(\mu,q)+(\nu,r)\leftrightarrow(\mu,s)+(\nu,t)$.

We now expand the field operators $\hat\psi_{\mu s}$ in the Wannier
basis for a 3-D lattice, such that
$\hat\psi_{\mu s}(x) = \sum_{i,n} \phi_{in}(x) \c_{in\mu s}$ with
spatial wavefunctions $\phi_{in}$ and fermionic annihilation operators
$\c_{in\mu s}$ indexed by lattice sites $i$ and motional states $n$.
Invoking the tight-binding approximation, we assume that the spatial
overlap integral in \eqref{eq:H_int_fields} is negligible unless all
wavefunctions are localized at the same lattice site; we discuss the
breakdown of this approximation and its consequences for our
low-energy effective theory in Appendix \ref{sec:error}.  The relevant
spatial overlap integral is then
\begin{align}
  K^{k\ell}_{mn}
  \equiv \int d^3x~ \phi_{im}^* \phi_{in}^* \phi_{i\ell} \phi_{ik},
  \label{eq:K_klmn}
\end{align}
which for a lattice with discrete translational invariance is
independent of the lattice site $i$.  The two-body interaction
Hamiltonian can be written in terms of this overlap integral as
\begin{align}
  H_{\t{int}}
  = \f12 \sum_{\substack{i,k,\ell,m,n\\q,r,s,t\\\mu,\nu}}
  K^{k\ell}_{mn} G^{qr}_{st}
  \c_{im\mu s}^\dag \c_{in\nu t}^\dag \c_{i\ell\nu r} \c_{ik\mu q}
  \equiv \f12 \sum K^{k\ell}_{mn} G^{qr}_{st}
  \c_{m\mu s}^\dag \c_{n\nu t}^\dag \c_{\ell\nu r} \c_{k\mu q},
  \label{eq:H_int}
\end{align}
where for brevity we will henceforth suppress the identical site index
($i$) on all operators, and implicitly sum over all free indices in a
summand (i.e.~indices which do not have a fixed value).  We may also
at times suppress motional state indices on the overlap integral
$K^{k\ell}_{mn}$, in which case the suppressed indices are implicitly
zero (corresponding to a motional ground state); i.e.
\begin{align}
  K^{\ell m}_n \equiv K^{\ell m}_{n,0},
  &&
  K^m_n \equiv K^{m,0}_{n,0},
  &&
  K_{mn} \equiv K^{0,0}_{mn},
  &&
  K_n \equiv K^{0,0}_{n,0},
  &&
  K \equiv K^{0,0}_{0,0}.
  \label{eq:K}
\end{align}
For simplicity, we will also generally work in a gauge for which all
two-body overlap integrals are real, such that
$K^{k\ell}_{mn}={K^{k\ell}_{mn}}^*=K^{mn}_{k\ell}$.  The existence of
such a gauge is guaranteed by the analytic properties of the Wannier
orbitals $\phi_{in}$ \cite{kohn1959analytic}.

We are interested in the ground and low-lying orbital eigenstates of
several interacting atoms on a single lattice site, as such states can
be readily prepared and coherently addressed in current experiments
with ultracold atoms \cite{goban2018emergence}.  Although these
experiments operate at temperatures sufficiently low for negligible
thermal occupation of excited motional states, simply neglecting
excited single-particle motional states fails to reproduce the
observed spectra of multiply-occupied lattice sites.  The reason for
this failure lies in the fact that the bare two-body Hamiltonian
$H_{\t{int}}$ in \eqref{eq:H_int} is not diagonal in the basis of
single-particle motional states.  Even at zero temperature, therefore,
the presence of excited motional states modifies many-body atomic
energy spectra when the magnitude of off-diagonal elements of
$H_{\t{int}}$ approach the scale of low-lying single-particle motional
state energy gaps.  To address this problem, we develop a low-energy
effective theory for ultracold two-level fermionic atoms on a lattice.
As our work is motivated by progress in experiments which initialize
all atoms in identical (i.e.~ground) orbital and motional states, to
simplify our theory we assume that any $N$ atoms on a single lattice
site occupy $N$ nuclear spin states.  Multiple occupation of a single
nuclear spin state on the same lattice site is initially forbidden by
fermionic statistics, and subsequently cannot be violated in the
absence of tunneling and hyperfine coupling.

Our low-energy effective theory exhibits SU($N$)-symmetric multi-body
interactions, such that the effective interaction Hamiltonian can be
written in the form
\begin{align}
  H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} H_M,
\end{align}
where $H_M$ is an $M$-body Hamiltonian, $I$ is the total nuclear spin
of each atom (e.g.~$I=9/2$ for ${}^{87}$Sr), and the sum terminates at
$2I+1$ because this is the largest number of atoms which may initially
occupy a single lattice site.  Focusing on the low-lying orbital state
excitations of this theory, we find that SU($N$) symmetry allows us to
express multi-body Hamiltonians in the simple form
\begin{align}
  H_M = \sum_{\abs{\set{\mu_j}}=M}
  \p{U_{M,\g} \prod_{\alpha=1}^M \n_{\mu_\alpha,\g}
    + U_{M,+} \n_{\mu_1,\e} \prod_{\alpha=2}^M \n_{\mu_\alpha,\g}
    + U_{M,-} \c_{\mu_1,\g}^\dag \c_{\mu_2,\e}^\dag
    \c_{\mu_2,\g} \c_{\mu_1,\e} \prod_{\alpha=3}^M \n_{\mu_\alpha,\g}},
  \label{eq:H_M_preview}
\end{align}
where $\c_{\mu s}\equiv\c_{0,\mu s}$ is a fermionic annihilation
operator for an atom in the motional ground state with nuclear spin
$\mu$ and orbital state $s$;
$\n_{\mu s}\equiv \c_{\mu s}^\dag \c_{\mu s}$ is a number operator;
and the sum is performed over all choices of nuclear spins $\mu_j$
with $j=1,2,\cdots,M$ for which all $\mu_j$ are distinct, or
equivalently all choices of $\mu_j$ for which the set $\set{\mu_j}$
contains $M$ elements, for a total of ${2I+1\choose M}\times\p{M!}$
nuclear spin combinations.  The coefficients $U_X$ can be expressed in
terms of the coupling constants of the bare two-body interaction
Hamiltonian (see Appendix \ref{sec:U_X}).  The key feature of the
$M$-body interactions in \eqref{eq:H_M_preview} is that they take a
form identical to the two-body interactions between motional
ground-state atoms, but with the addition of $M-2$ spectator atoms.
This form is a direct consequence of the SU($N$) symmetry of
underlying two-body interactions.

@@

As the bare two-body Hamiltonian $H_{\t{int}}$ is not diagonal with
respect to single-particle motional states, the problem of determining
many-body spectra and eigenstates nominally involves all motional
degrees of freedom of the atoms involved.  If we are interested in
addressing internal atomic states at zero temperature, however, in
principle we need only to consider the motional ground state
associated with each many-body internal atomic state.  We can then
ignore all excited many-body motional states, as they will be neither
thermally occupied nor externally interrogated.  This procedure makes
the many-body problem more tractable by greatly reducing the
dimensionality of our the relevant Hilbert space.

@@

order, we introduce an expanded coupling symbol

@@

indices, the expanded coupling symbol allows us to collect together
diagrams which have the same graph topology, but represent different

@@

associated minus sign if they contain only one excited virtual state,
and a factor of $1/2$ if they contain a virtual ground state.

@@

in a contribution to the effective Hamiltonian that is equal to

@@

the second and third of these diagrams pick up a factor of $-1/2$, so
the sum over disconnected diagrams vanishes.
